---
title: "Tutorial for merging satellite-based precipitation datasets with ground observations using RFmerge (>=0.3-0)"
author:
- "Mauricio Zambrano-Bigiarini^[mauricio.zambrano@ufrontera.cl]"
- "Oscar M. Baez-Villanueva^[oscar.baezvillanueva@ugent.be]"
- "Juan Diego Giraldo-Osorio"
- Ian McNamara
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
  pdf_document:
    number_sections: true
vignette: |
  %\VignetteIndexEntry{Tutorial for merging satellite-based precipitation datasets with ground observations using RFmerge (>=0.3-0)} 
  %\VignetteKeyword{hydrology}
  %\VignetteKeyword{hydrological modelling}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown} 
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# About
This vignette is its based in a similar vignette, developed for `RFmerge` version 0.1-6 or earlier, which worked only with the superseded `raster` package. 

This vignette, based on the `terra` package, describes a basic application of the `RFmerge` function to create an improved precipitation dataset by combining two different satellite-based precipitation products with ground-based observations and user-selected covariates (i.e., digital elevation model and Euclidean distances to rain gauge stations).

We use as case study Valparaiso (Chile), as an example of how to generate this product at daily temporal scale and at $0.05^\circ$ spatial resolution, from January to August 1983. This example requires the following data from the user: 

  i) time series of rainfall observations, 
    
  ii) metadata describing the spatial coordinates of the rain gauges,
    
  iii) the Climate Hazards Group InfraRed Precipitation with Station data version 2.0 (CHIRPSv2), 
    
  iv) the Precipitation Estimation from Remotely Sensed Information using Artificial Neural Networks - Climate Data Record (PERSIANN-CDR), and
    
  v) the Shuttle Radar Topography Mission (SRTM-v4) digital elevation model (DEM). 
  
In addition, Euclidean distances are also used as covariates. They are automatically computed within `RFmerge` as the Euclidean distances from each rain gauge station to every grid-cell within the study area.



# Citation

If you find this tutorial useful, please cite it as Zambrano-Bigiarini et al. (2026):

> Zambrano-Bigiarini, M.; Baez-Villanueva, O.M.; Giraldo-Osorio, J.D. (2026). Tutorial for merging satellite-based precipitation datasets with ground observations using RFmerge (>=0.3-0). doi:10.5281/zenodo.20061104.

The theoretical description of the merging algorithm is described in the following RSE article:

> Baez-Villanueva, O. M.; Zambrano-Bigiarini, M.; Beck, H.; McNamara, I.; Ribbe, L.; Nauditt, A.; Birkel, C.; Verbist, K.; Giraldo-Osorio, J.D.; Thinh, N.X. (2020). [RF-MEP: a novel Random Forest method for merging gridded precipitation products and ground-based measurements](https://doi.org/10.1016/j.rse.2019.111606), Remote Sensing of Environment, 239, 111610. doi:10.1016/j.rse.2019.111606.

Last, but not least, you should cite the `RFmerge` R package as follows:

> Zambrano-Bigiarini, M.; Baez-Villanueva, O.M., Giraldo-Osorio, J. (2026). RFmerge: Merging of Satellite Datasets with Ground Observations using Random Forests. R package version 0.3-0. URL: https://hzambran.github.io/RFmerge/. doi:10.32614/CRAN.package.RFmerge.



# Required `RFmerge` version

This tutorial was developed for `RFmerge >= 0.3-0`, which is the minimum version where `RFmerge` is based on the [terra](https://cran.r-project.org/package=terra) package.

All the previous `RFmerge` versions until 0.1-6 were based on the superseded [raster](https://cran.r-project.org/package=raster) package. `RFmerge`  0.1-6 was removed from CRAN in 2023-02-25, because issues related to old spatial packages (e.g., rgdal) were not solved in time. 




# Installation

Install the latest stable version (from CRAN):

```{r Install_Stable, eval = FALSE}
install.packages("RFmerge")
```

Alternatively, you can also try the under-development version from [Github](https://github.com/hzambran/RFmerge):
```{r Install_Unstable, eval = FALSE}
if (!require(devtools)) install.packages("devtools")
library(devtools)
install_github("hzambran/RFmerge")
```

\newpage
# Setting up the environment

1. Load other packages that will be used in this analysis:

```{r Loading_other_pks, eval = TRUE, message=FALSE}
library(zoo)
library(terra)
```

2. Load the `RFmerge` package, which contains the main function used in the analysis and required datasets:

```{r LoadingRFmerge, eval = TRUE, message=FALSE}
library(RFmerge)
```

# Loading input data

First, daily time series of rainfall observations in 34 rain gauges located in Valparaiso will be used for this example, from *1983-01-01* to *1983-08-31*, which are available in the **ValparaisoPPts** dataset provided in the `RFmerge` package (for your own application, this dataset might be read from a CSV file or a `zoo` file). In addition, the **ValparaisoPPgis** dataset contains information about the IDs and spatial coordinates of each one of the rain gauges (for your own application, this dataset might be read from a CSV file). Finally, **ValparaisoSHP** stores the `sf` polygon defining the outer borders of the study area (for your own application, this dataset might be read from an ESRI shapefile).
  
   
```{r Loading_GroundObservarions, eval = TRUE}
data(ValparaisoPPts)    
data(ValparaisoPPgis) 
```


   
```{r Loading_ValparaisoShapefile, eval = TRUE}
ValparaisoSHP.fname <- system.file("extdata/ValparaisoSHP.shp",package="RFmerge")
ValparaisoSHP       <- terra::vect(ValparaisoSHP.fname)
```


Secondly, we need to load the satellite-based precipitation datasets and other covariates. For this example, CHIRPSv2 (Funk et al., 2015) and PERSIANN-CDR (Ashouri et al.,2015), at a spatial resolution of $0.05^\circ$, are used as dynamic covariates (in this context, *dynamic* means time-varying covariates). The Digital Elevation Model SRTM-v4 (DEM), also with a spatial resolution of $0.05^\circ$, is used as a static covariate to account for the impact of elevation on precipitation(in this context, *static* means that this covariate does not change in time).

   
```{r LoadingSatelliteData, eval = TRUE}
chirps.fname   <- system.file("extdata/CHIRPS5km.tif"       ,package="RFmerge")
prsnncdr.fname <- system.file("extdata/PERSIANNcdr5km.tif"  ,package="RFmerge")
dem.fname      <- system.file("extdata/ValparaisoDEM5km.tif",package="RFmerge")

CHIRPS5km        <- terra::rast(chirps.fname)
PERSIANNcdr5km   <- terra::rast(prsnncdr.fname)
ValparaisoDEM5km <- terra::rast(dem.fname)   
```


# Basic exploratory data analysis

The two multi-band geotiff files provided in the `RFmerge` package do not store the date of the precipitation estimate as name of the corresponding layer. Therefore, before any exploratory analysis, we would give meaningful names to each band (layer) in `CHIRPS5km` and `PERSIANNcdr5km`:

```{r GivingMeaninfulNamesToLayers, eval = TRUE}
#ldates                <- hydroTSM::dip("1983-01-01", "1983-08-31")
ldates                <- seq(from=as.Date("1983-01-01"), to=as.Date("1983-08-31"), by="days")
names(CHIRPS5km)      <- ldates
names(PERSIANNcdr5km) <- ldates  
```

Then, we want to visualise the first six rows of the spatial metadata:

```{r ExploringMetadata}
head(ValparaisoPPgis)
```


Plotting the daily precipitation time series for the first station (code: *P5101005*).

```{r PlottingPts, fig.width = 7, fig.height = 3.5, fig.align = "center", message=FALSE}
main <- paste("Daily precipitation for the station", ValparaisoPPgis$Code[1])
ylab <- "Precipitation [mm]"
x.ts <- ValparaisoPPts[,1]

#hydroTSM::hydroplot(x.ts, pfreq="o", main=main, ylab= ylab)
plot(x.ts, main=main, ylab= ylab, col="blue")
grid()
```



Plotting the accumulated precipitation estimates for the first eight months of 1983 from CHIRPS and PERSIANN-CDR, and overlying the boundaries of the study area (only its first attribute):

```{r PlotingTotalP, fig.width= 4, fig.height = 4, fig.align = "center", message=FALSE}
chirps.total   <- sum(CHIRPS5km, na.rm= FALSE)
persiann.total <- sum(PERSIANNcdr5km, na.rm= FALSE)


plot(chirps.total, main = "CHIRPSv2 [Jan - Aug] ", 
     xlab = "Longitude", ylab = "Latitude",
     fun=function() lines(ValparaisoSHP))

plot(persiann.total, main = "PERSIANN-CDR [Jan - Aug]", 
     xlab = "Longitude", ylab = "Latitude",
     fun=function() lines(ValparaisoSHP))
```


# Preparing input data

## Spatial metadata

In order to easily plot the spatial information about the in situ gauging stations stored in `ValparaisoPPgis`, we first need to convert it into a `SpatVector`, using the latitude and longitude fields stored in the `lat` and `lon` columns of the `ValparaisoPPgis` data.frame object:

```{r SpatialMetadata}
ValparaisoPPgis.vec <- terra::vect(ValparaisoPPgis, geom=c("lon", "lat"), crs="epsg:4326")
```


Now, we can plot the DEM along with the locations of the rain gauge stations that will be used in the computation of RF-MEP, and overlying the boundaries of the study area and the stations (only their first attribute):

```{r PlottingDEMwithObs, fig.width = 4, fig.height = 4, fig.align = "center"}
plot(ValparaisoDEM5km, main="SRTM-v4", xlab="Longitude", ylab="Latitude", 
     col=terrain.colors(255),
     fun=function() {lines(ValparaisoSHP) ; 
                     points(ValparaisoPPgis.vec, pch=16, col="black") 
                    }
    )
```

## Verification of covariates

Note that for this example we want to produce a merged precipitation product only from January 1st to August 31th in 1983, i.e., 243 days; therefore, the precipitation products used as covariates have 243 layers each (one for each day in the time period used ofr the analysis), which is the **same number of rows** of the `ValparaisoPPts` object.

```{r nLayersVerification}
terra::nlyr(CHIRPS5km)
( terra::nlyr(CHIRPS5km) == terra::nlyr(PERSIANNcdr5km) )
( terra::nlyr(CHIRPS5km) == nrow(ValparaisoPPts) )
```

Now, we have to verify that the precipitation products and the DEM have the same extent, number of rows and columns, projection, resolution, and origin:

```{r SGeometryVerification}
terra::compareGeom(CHIRPS5km, PERSIANNcdr5km, ValparaisoDEM5km)
```


### Optional: Reprojection into another CRS

If you would like to test the use of Euclidean distances as covariates in `RFmerge` (using the argument `ED=TRUE`) you might want to re-project your covariates and in situ stations from geographical coordinates into a projected coordinate system. However, for the computation of distances in `terra`, this package uses the `distance` function, which explicitly mentions: 


> *The distance is always expressed in meter if the coordinate reference system is longitude/latitude, and in map units otherwise. Map units are typically meter, but inspect crs(x) if in doubt*.

> *Results are more precise, sometimes much more precise, when using longitude/latitude rather than a planar coordinate reference system, as these distort distance*. 

Therefore, in contrast to the recommendation given by the previous version of this package (the one based on the raster package), **we are not going to re-project any input dataset**.





# Running `RFmerge` using geographical coordinates

## Covariates

To run `RFmerge` we first need to create the covariates object required by this function. For doing this, we will create a list object with the different covariates. 

```{r CovariatesCreation}
covariates<- list(chirps=CHIRPS5km, persianncdr=PERSIANNcdr5km, 
                  dem=ValparaisoDEM5km)
```

> Please note that **the order and name of the covariates in the list is not important**.


## Setup

Now, you need to decide how much of you in situ observations are going to be used for training the random forest model and how much of them are going to be left for validation. 

In this example, we are going to use 80% of the in situ gauging stations for training and the remaining 20% for validation:

```{r RFmergeWithoutParallelisationGEO, eval = TRUE}
geoNoPar.drty.out <- file.path(tempdir(), "RFmergeResults_geographical_noparallel")
system.time(rfmep.NoPar.geo <- RFmerge(x=ValparaisoPPts, 
                                      cov=covariates,
                                      metadata=ValparaisoPPgis, 
                                      id="Code", lat="lat", lon="lon", 
                                      mask=ValparaisoSHP, 
                                      training=0.8, 
                                      seed=1000,
                                      write2disk=TRUE, 
                                      drty.out=geoNoPar.drty.out
                                      ) 
            )
```

In the previous example, we used the argument `seed=1000` to be able to obtain reproducible results with random forest.

If you want the resulting merged files to be written into disk for further inspection, you need to set the argument `write2disk=TRUE` and define the output directory (`drty.out`) before running `RFmerge`.


# Running `RFmerge` using a projected reference system

## Re-projection

To evaluate how large are the differences between the results obtained with input datasets in geographical coordinates vs  results obtained with input datasets in projected coordinates, we are going to create input datasets in projected coordinates, using WGS 84 / UTM zone 19S (EPSG:32719) as reference system:

First, we re-project all the gridded covariates:

```{r ReprojectingRasters}
utmz19s.p4s <- "epsg:32719" # WGS 84 / UTM zone 19S

CHIRPS5km.utm        <- terra::project(x=CHIRPS5km       , y=utmz19s.p4s)
PERSIANNcdr5km.utm   <- terra::project(x=PERSIANNcdr5km  , y=utmz19s.p4s)
ValparaisoDEM5km.utm <- terra::project(x=ValparaisoDEM5km, y=utmz19s.p4s)
```

Second, we re-project the in situ gauging stations:

```{r ReprojectingMetadata}
ValparaisoPPgis.vec.utm <- terra::project(x=ValparaisoPPgis.vec, y=utmz19s.p4s)
```



Third, we re-project the polygon with the boundaries used to define the study area:

```{r ReprojectingSHP}
ValparaisoSHP.utm <- terra::project(ValparaisoSHP, y=utmz19s.p4s)
```


Fourth, we create a new data.frame with the expected metadata( i.e., at least, ID, lat, lon) in the re-projected coordinates:

```{r FinalMetadata}
id        <- ValparaisoPPgis.vec.utm[["Code"]][,1]
st.coords <- terra::crds(ValparaisoPPgis.vec.utm)
lon       <- st.coords[, "x"]
lat       <- st.coords[, "y"]

ValparaisoPPgis.utm <- data.frame(Code=id, lon=lon, lat=lat)
```

## Covariates

Now, we are going to create the covariates object to be used in the `RFmerge` function. For doing this, we will create a list object with the different covariates. 

```{r CovariatesCreationUTM}
covariates.utm <- list(chirps=CHIRPS5km.utm, persianncdr=PERSIANNcdr5km.utm, 
                       dem=ValparaisoDEM5km.utm)
```

> Please note that **the order and name of the covariates in the list is not important**.


## Setup

To obtain results comparable to the ones obtained in the example run with input variables in geographical coordinates, we are going to use the same 80% of the in situ gauging stations for training and the same remaining 20% for validation:

```{r RFmergeWithoutParallelisationUTM, eval = TRUE}
utmNoPar.drty.out <- file.path(tempdir(), "RFmergeResults_utm_noparallel")
system.time(rfmep.NoPar.utm <- RFmerge(x=ValparaisoPPts, 
                                      cov=covariates.utm,
                                      metadata=ValparaisoPPgis.utm, 
                                      id="Code", lat="lat", lon="lon", 
                                      mask=ValparaisoSHP.utm, 
                                      training=0.8, 
                                      seed=1000,
                                      write2disk=TRUE, 
                                      drty.out=utmNoPar.drty.out
                                     ) 
            )
```


# Running `RFmerge` using the `parallel` argument

first of all, it is necessary to mention that one of the main advantages of moving `RFmerge` from the superseded `raster` package to its newest version `terra`, is an enormous gain of speed in the computations. 

However, if for your particular application you want to take advantage of your multi-core machine or from a existing network cluster, you can do it using the `parallel` argument included in the `RFmerge` function.

First, we are going to define the desired number of cores/nodes to be used in the parallel computations:

```{r NcoresUserDefined, eval = TRUE}
ncores.nmax <- 10 # maximum amount of cores the user wants to use
```

Now, we are going to detect if your OS is Windows or GNU/Linux, and setting the 'parallel' argument accordingly:

```{r OnWindows, eval = TRUE}
onWin <- ( (R.version$os=="mingw32") | (R.version$os=="mingw64") )
ifelse(onWin, parallel <- "parallelWin", parallel <- "parallel")
```

Then, we proceed to make a safe selection of the number of cores to be used in the computations, by comparing the user-defined number (`ncores.nmax`) with the maximum number of cores/nodes available in you environment (`parallel::detectCores()`), and avoiding an invalid number of nodes when `detectCores()` returns NA or 1:

```{r NcoresDetection, eval = TRUE}
( ncores.available <- parallel::detectCores() )
if (is.na(ncores.available)) ncores.available <- 2
( par.nnodes <- max(1, min(ncores.available - 1, ncores.nmax) ) )
```

Now we are ready to use `RFmerge` using parallelisation in a GNU/Linux or macOS machine, with a maximum number of nodes/cores to be used equal to `ncores.nmax`:

```{r RFmergeWithLinuxParallelisation, eval = TRUE}
geoPar.drty.out <- file.path(tempdir(), "RFmergeResults_geographical_parallel")
system.time(rfmep.Par.geo <- RFmerge(x=ValparaisoPPts, 
                                    cov=covariates,
                                    metadata=ValparaisoPPgis, 
                                    id="Code", lat="lat", lon="lon", 
                                    mask=ValparaisoSHP, 
                                    training=0.8, 
                                    seed=1000,
                                    write2disk=TRUE,  
                                    drty.out=geoPar.drty.out, 
                                    parallel=parallel, par.nnodes=par.nnodes
                                   )
            )
```


# `RFmerge` outputs

If `RFmerge` run without problems, and `write2disk=TRUE`, the following output files will be stored in your user-defined `drty.out` directory: 

  i) the rain gauge stations used as training and evaluation datasets; and 
 
  ii) the final merged product (individual *GeoTiff* files).
  
  
The aforementioned output objects will be stored within `drty.out` as follows:


- `Ground_based_data/Training/`: In this directory, time series and metadata used to train RF-MEP will be stored as a `zoo` file (`Training_ts.txt`), and a `text` file (`Training_metadata.txt`), respectively. 

- `Ground_based_data/Evaluation/`: This directory will store time series and metadata not used in training the Random Forest model, but that are available for evaluating the performance of the results in an independent evaluation dataset. The `Evaluation_ts.txt` and `Evaluation_metadata.tx` files are stored as `zoo` and  `CSV` files, respectively)

- `RF-MEP/`: This directory will store the individual *GeoTiff* files produced by the RF-MEP algorithm, using the same spatial resolution as the selected covariates.


# Evaluation of `RFmerge` using geographical coordinates

## Reading `RFmerge` outputs

To evaluate the outputs obtained with `RFmerge` in **geographical coordinates** we will use the evaluation dataset of rain gauge observations not used in the merging procedure. 

For this purpose, we will create a SpatRaster object from the obtained merged product and also import the time series and metadata from the evaluation set of gauging stations:


```{r Evaluation_geoNoPar.drty.out, eval= TRUE}
ts.path       <- file.path(geoNoPar.drty.out, "Ground_based_data/Evaluation/Evaluation_ts.txt")
metadata.path <- file.path(geoNoPar.drty.out, "Ground_based_data/Evaluation/Evaluation_metadata.txt")

eval.ts  <- read.zoo(ts.path, header = TRUE)
eval.gis <- read.csv(metadata.path)

# promoting 'eval.gis' into a spatial object (in order to be plotted)
( eval.gis.vec <- terra::vect(eval.gis, geom=c("lon", "lat")) )
```
## Basic plotting of `RFmerge` outputs for a single day

First, we define the colours and categories used for the subsequent plots of the output merged dataset:

```{r BreaksAndColoursDefinition, eval= TRUE}
minmax(rfmep.NoPar.geo[[11]])                       # range of values to be plotted

breaks <- c(0, 2, 4, 6, 8, 10, 12, 15, 20)         # 8 categories for the precipitation values
colors <- RColorBrewer::brewer.pal(n=8, "YlGnBu")  # 8 colours
```

Basic visualisation of one day of precipitation obtained with RF-MEP (in this case, 1983-01-11), and overlying the boundaries of the study area and the gauging stations used for evaluation:

```{r BasicOutputVis, fig.width=4, fig.height=4, fig.align="center", eval=TRUE}
plot(rfmep.NoPar.geo[[11]],
     main="RF-MEP precipitation for 1983-01-11", 
     xlab="Longitude", ylab="Latitude", 
     breaks=breaks, col=colors
     )
lines(ValparaisoSHP, col="black")
points(eval.gis.vec, pch = 18)
```

## Advanced plotting of `RFmerge` outputs for a single day


Now, we are going to use a more advanced visualization of one day of precipitation obtained with RF-MEP (in this case, 1983-01-11). In order to do that, we first map the precipitation values at the evaluation gauging stations with the previously defined colour palette:

```{r AdvancedOutputVisColours, fig.width=4, fig.height=4, fig.align="center", eval=TRUE}
stations.vals   <- as.numeric(eval.ts[11,])
stations.colors <- cut(stations.vals, breaks=breaks, 
                       labels=colors, include.lowest=TRUE)
stations.colors <- as.character(stations.colors) # Convert factors to colour strings
```

Now, we overly the boundaries of the study area, and colouring the gauging stations used for evaluation with the same colours used for the output merged dataset, in order to quickly asses how far or how close is the output merged dataset to the gauging stations used for evaluation only.

```{r AdvancedOutputVisPlotting, fig.width=5.2, fig.height=5.2, fig.align="center", eval=TRUE}
plot(rfmep.NoPar.geo[[11]], 
     main="RF-MEP precipitation for 1983-01-11", 
     xlab="Longitude", ylab="Latitude", 
     breaks=breaks, col=colors,
     plg = list(title = "P, [mm]")
     )
lines(ValparaisoSHP, col="black")

# Adding labels from a specific column in the SpatVector
# -) labels: The name of the column in your SpatVector containing the text 
#            you want to display.
# -) halo  : Set to TRUE to add a buffer around the text, making it more 
#            readable against complex raster backgrounds.
# -) pos   : Determines the position relative to the point 
#            (1=below, 2=left, 3=above, 4=right).
points(eval.gis.vec, pch = 23, # Diamond with fill and border
       bg = stations.colors,      # Fill colour
       col = "black",             # Border (outline) colour
       cex=0.9)
text(eval.gis.vec, labels="Code", halo=TRUE, pos=3, cex=0.4, col="black")
text(eval.gis.vec, label=paste(stations.vals, "mm"), 
     halo=TRUE, pos=1, cex=0.4, col="black")
```

## Plotting the total amount of precipitation

The total amount of precipitation over Valparaiso for January to August 1983 according to RF-MEP, and overlying the boundaries of the study area (only its first attribute)::

```{r TotalP, fig.width = 4, fig.height = 4, fig.align = "center", eval= TRUE}
rfmep.total <- sum(rfmep.NoPar.geo, na.rm= FALSE)

plot(rfmep.total, main = "RF-MEP [Jan - Aug]", 
     xlab = "Longitude", ylab = "Latitude")
lines(ValparaisoSHP, lwd=1.5)
```


## Numeric assessment of `RFmerge' effectiveness

We will evaluate the effectiveness of the merged product obtained with `RFmerge` in **geographical coordinates**, hereafter RF-MEP,  and the two precipitation products used as covariates, against the in situ rain gauge data from the evaluation set. 

To extract the gridded precipitation values at the rain gauge locations, we will use the `extract` function from the `terra` package:

```{r EvaluationExtraction, eval= TRUE}
eval.gis.vec <- vect(eval.gis, geom = c("lon", "lat"))

rfmep.NoPar.geo.ts <- t(terra::extract(rfmep.NoPar.geo, eval.gis.vec))
chirps.ts         <- t(terra::extract(CHIRPS5km     , eval.gis.vec))
persiann.ts       <- t(terra::extract(PERSIANNcdr5km, eval.gis.vec))
```

The output of the `extract` function adds a first row with the ID of the points used during the extraction. This additional row must be removed before using those time series for comparison with any of the gridded precipitation datasets:

```{r RemovalOfUnnecessaryRow, eval= TRUE}
rfmep.NoPar.geo.ts <- rfmep.NoPar.geo.ts[-1, ]
chirps.ts         <- chirps.ts[-1, ]
persiann.ts       <- persiann.ts[-1, ]
```

To evaluate the performance of RF-MEP product obtained with `RFmerge` and the products used in its computation, considering the original geographical coordinates, we are going to use the Nash-Sutcliffe efficiency (NSE), as provided by the [hydroGOF](https://mzb.cl/hydroGOF/) R package. NSE ranges from negative infinity to one, with an optimal value of 1.


```{r EvaluationComputation, eval= TRUE}
sres       <- list(chirps.ts, persiann.ts, rfmep.NoPar.geo.ts)
nsres      <- length(sres)
nstations  <- ncol(eval.ts)
tmp        <- rep(NA, nstations)
nse.table.NoPar.geo <- data.frame(ID=eval.gis[["Code"]], 
                                 CHIRPS=tmp, 
                                 PERSIANN_CDR=tmp, 
                                 RF_MEP=tmp)

# Computing the NSE between the observed rainfall measured in each one of the 
# rain gauges of the training dataset and CHIRPSv2, PERSIANN-CDR, 
# the merged product `rfmep.NoPar.geo.ts`:
for (i in 1:nsres) {
  ldates <- time(eval.ts)
  lsim   <- zoo(sres[[i]], ldates)
  nse.table.NoPar.geo[, (i+1)] <- round( hydroGOF::NSE(sim= lsim, obs= eval.ts, 
                                                     na.rm=TRUE), 3 )
} # FOR end

# Showing the final table with the performance of all products
nse.table.NoPar.geo
```

## Graphical assessment of `RFmerge' effectiveness

Finally, we create a boxplot comparing the performance, in terms of *NSE*, of the merged product in comparison to the original SREs used as covariates:

```{r EvaluationPlotting, fig.width=5, fig.height=5, fig.align="center", eval=TRUE}

# Boxplot with a graphical comparison
sres.cols <- c("powderblue", "palegoldenrod", "mediumseagreen")
boxplot(nse.table.NoPar.geo[,2:4], 
        main = "Daily NSE evaluation for Jan - Aug 1983 \n using geographical coordinates",
        xlab = "P products", ylab = "NSE'", ylim = c(-0.5, 1), # horizontal=TRUE,
        col=sres.cols)
legend("topleft", legend=c("CHIRPS", "PERSIANN-CDR", "RF-MEP"), 
       col=sres.cols, pch=15, cex=1.5, bty="n")
grid()
```




# Evaluation of `RFmerge` using projected coordinates

## Reading `RFmerge` outputs

To evaluate the outputs obtained with `RFmerge` in **projected (UTM) coordinates** we will use the evaluation dataset of rain gauge observations not used in the merging procedure. 

For this purpose, we will create a SpatRaster object from the obtained merged product and also import the time series and metadata from the evaluation set of gauging stations:


```{r Evaluation_geoNoPar.drty.out_UTM, eval= TRUE}
ts.path       <- file.path(utmNoPar.drty.out, "Ground_based_data/Evaluation/Evaluation_ts.txt")
metadata.path <- file.path(utmNoPar.drty.out, "Ground_based_data/Evaluation/Evaluation_metadata.txt")

eval.ts      <- read.zoo(ts.path, header = TRUE)
eval.gis.utm <- read.csv(metadata.path)

# promoting 'eval.gis' into a spatial object (in order to be plotted)
( eval.gis.vec.utm <- terra::vect(eval.gis.utm, geom=c("lon", "lat"), crs=utmz19s.p4s ) )
```

## Basic plotting of `RFmerge` outputs for a single day

First, we define the colours and categories used for the subsequent plots of the output merged dataset:

```{r BreaksAndColoursDefinition_UTM, eval= TRUE}
minmax(rfmep.NoPar.utm[[11]])                       # range of values to be plotted

breaks <- c(0, 2, 4, 6, 8, 10, 12, 15, 20)         # 8 categories for the precipitation values
colors <- RColorBrewer::brewer.pal(n=8, "YlGnBu")  # 8 colours
```

Basic visualisation of one day of precipitation obtained with RF-MEP (in this case, 1983-01-11), and overlying the boundaries of the study area and the gauging stations used for evaluation:

```{r BasicOutputVis_UTM, fig.width=4, fig.height=4, fig.align="center", eval=TRUE}
plot(rfmep.NoPar.utm[[11]],
     main="RF-MEP precipitation for 1983-01-11", 
     xlab="Longitude", ylab="Latitude", 
     breaks=breaks, col=colors
     )
lines(ValparaisoSHP.utm, col="black")
points(eval.gis.vec.utm, pch = 18)
```


## Advanced plotting of `RFmerge` outputs for a single day

Now, we are going to use a more advanced visualization of one day of precipitation obtained with RF-MEP (in this case, 1983-01-11). In order to do that, we first map the precipitation values at the evaluation gauging stations with the previously defined colour palette:

```{r AdvancedOutputVisColours_UTM, fig.width=4, fig.height=4, fig.align="center", eval=TRUE}
stations.vals   <- as.numeric(eval.ts[11,])
stations.colors <- cut(stations.vals, breaks=breaks, 
                       labels=colors, include.lowest=TRUE)
stations.colors <- as.character(stations.colors) # Convert factors to colour strings
```

Now, we overly the boundaries of the study area, and colouring the gauging stations used for evaluation with the same colours used for the output merged dataset, in order to quickly asses how far or how close is the output merged dataset to the gauging stations used for evaluation only.

```{r AdvancedOutputVisPlotting_UTM, fig.width=5.2, fig.height=5.2, fig.align="center", eval=TRUE}
plot(rfmep.NoPar.utm[[11]], 
     main="RF-MEP precipitation for 1983-01-11", 
     xlab="Longitude", ylab="Latitude", 
     breaks=breaks, col=colors,
     plg = list(title = "P, [mm]")
     )
lines(ValparaisoSHP.utm, col="black")

# Adding labels from a specific column in the SpatVector
# -) labels: The name of the column in your SpatVector containing the text 
#            you want to display.
# -) halo  : Set to TRUE to add a buffer around the text, making it more 
#            readable against complex raster backgrounds.
# -) pos   : Determines the position relative to the point 
#            (1=below, 2=left, 3=above, 4=right).
points(eval.gis.vec.utm, pch = 23, # Diamond with fill and border
       bg = stations.colors,      # Fill colour
       col = "black",             # Border (outline) colour
       cex=0.9)
text(eval.gis.vec.utm, labels="Code", halo=TRUE, pos=3, cex=0.4, col="black")
text(eval.gis.vec.utm, label=paste(stations.vals, "mm"), 
     halo=TRUE, pos=1, cex=0.4, col="black")
```

## Plotting the total amount of precipitation

The total amount of precipitation over Valparaiso for January to August 1983 according to RF-MEP, and overlying the boundaries of the study area (only its first attribute)::

```{r TotalP_UTM, fig.width = 4, fig.height = 4, fig.align = "center", eval= TRUE}
rfmep.total.utm <- sum(rfmep.NoPar.utm, na.rm= FALSE)

plot(rfmep.total.utm, main = "RF-MEP [Jan - Aug]", 
     xlab = "Longitude", ylab = "Latitude")
lines(ValparaisoSHP.utm, lwd=1.5)
```


## Numeric assessment of `RFmerge' effectiveness

We will evaluate the effectiveness of the merged product obtained with `RFmerge` in **geographical coordinates**, hereafter RF-MEP,  and the two precipitation products used as covariates, against the in situ rain gauge data from the evaluation set. 

To extract the gridded precipitation values at the rain gauge locations, we will use the `extract` function from the `terra` package:

```{r EvaluationExtraction_UTM, eval= TRUE}
rfmep.NoPar.utm.ts <- t(terra::extract(rfmep.NoPar.utm   , eval.gis.vec.utm))
chirps.ts.utm      <- t(terra::extract(CHIRPS5km.utm     , eval.gis.vec.utm))
persiann.ts.utm    <- t(terra::extract(PERSIANNcdr5km.utm, eval.gis.vec.utm))
```

The output of the `extract` function adds a first row with the ID of the points used during the extraction. This additional row must be removed before using those time series for comparison with any of the gridded precipitation datasets:

```{r RemovalOfUnnecessaryRow_UTM, eval= TRUE}
rfmep.NoPar.utm.ts <- rfmep.NoPar.utm.ts[-1, ]
chirps.ts.utm     <- chirps.ts.utm[-1, ]
persiann.ts.utm   <- persiann.ts.utm[-1, ]
```

To evaluate the performance of RF-MEP product obtained with `RFmerge` and the products used in its computation, considering the original geographical coordinates, we are going to use the Nash-Sutcliffe efficiency (NSE), as provided by the [hydroGOF](https://mzb.cl/hydroGOF/) R package. NSE ranges from negative infinity to one, with an optimal value of 1.


```{r EvaluationComputation_UTM, eval= TRUE}
sres       <- list(chirps.ts.utm, persiann.ts.utm, rfmep.NoPar.utm.ts)
nsres      <- length(sres)
nstations  <- ncol(eval.ts)
tmp        <- rep(NA, nstations)
nse.table.NoPar.utm <- data.frame(ID=eval.gis.utm[["Code"]], 
                                 CHIRPS=tmp, 
                                 PERSIANN_CDR=tmp, 
                                 RF_MEP=tmp)

# Computing the NSE between the observed rainfall measured in each one of the 
# rain gauges of the training dataset and CHIRPSv2, PERSIANN-CDR, the merged 
# product `rfmep.NoPar.utm.ts`:
for (i in 1:nsres) {
  ldates <- time(eval.ts)
  lsim   <- zoo(sres[[i]], ldates)
  nse.table.NoPar.utm[, (i+1)] <- round( hydroGOF::NSE(sim= lsim, obs= eval.ts, 
                                                     na.rm=TRUE), 3 )
} # FOR end

# Showing the final table with the performance of all products
nse.table.NoPar.utm
```

## Graphical assessment of `RFmerge' effectiveness

Finally, we create a boxplot comparing the performance, in terms of *NSE*, of the merged product in comparison to the original SREs used as covariates:

```{r EvaluationPlotting_UTM, fig.width=5, fig.height=5, fig.align="center", eval=TRUE}

# Boxplot with a graphical comparison
sres.cols <- c("powderblue", "palegoldenrod", "mediumseagreen")
boxplot(nse.table.NoPar.utm[,2:4], 
        main = "Daily NSE evaluation for Jan - Aug 1983 \n using geographical coordinates",
        xlab = "P products", ylab = "NSE'", ylim = c(-0.5, 1), # horizontal=TRUE,
        col=sres.cols)
legend("topleft", legend=c("CHIRPS", "PERSIANN-CDR", "RF-MEP"), 
       col=sres.cols, pch=15, cex=1.5, bty="n")
grid()
```





# Software details

```{r hydroGOF_RColorBrewer, eval = TRUE, include = FALSE}
library(hydroGOF)
library(RColorBrewer)
```

This tutorial was built under: 

```{r SoftwareDetails, echo=FALSE, eval = TRUE}
sessionInfo()$platform
sessionInfo()$R.version$version.string 
paste("RFmerge", sessionInfo()$otherPkgs$RFmerge$Version)
paste("terra", sessionInfo()$otherPkgs$terra$Version)
paste("hydroGOF", sessionInfo()$otherPkgs$hydroGOF$Version)
paste("RColorBrewer", sessionInfo()$otherPkgs$RColorBrewer$Version)
```

\newpage

# References

1. Ashouri, H., Hsu, K.-L., Sorooshian, S., Braithwaite, D. K., Knapp, K. R., Cecil, L. D., Nelson, B. R., and Prat, O. P. (2015). PERSIANN-CDR:
Daily Precipitation Climate Data Record from Multisatellite Observations for Hydrological and Climate Studies, Bulletin of the American
Meteorological Society, 96, 69–83, doi:10.1175/BAMS-D-13-00068.1.

2. Baez-Villanueva, O. M.; Zambrano-Bigiarini, M.; Beck, H.; McNamara, I.; Ribbe, L.; Nauditt, A.; Birkel, C.; Verbist, K.; Giraldo-Osorio, J.D.; Thinh, N.X. (2020). RF-MEP: a novel Random Forest method for merging gridded precipitation products and ground-based measurements, Remote Sensing of Environment, 239, 111610. doi:10.1016/j.rse.2019.111606. https://doi.org/10.1016/j.rse.2019.111606.

3. Funk, C., Peterson, P., Landsfeld, M., Pedreros, D., Verdin, J., Shukla, S., Husak, G., Rowland, J., Harrison, L., Hoell, A., and Michaelsen,
J. (2015) The climate hazards infrared precipitation with stations-a new environmental record for monitoring extremes, Sci Data, 2, 150 066, doi:10.1038/sdata.2015.66.

4. Hengl, T., Nussbaum, M., Wright, M. N., Heuvelink, G. B., & Gr\"{a}ler, B. (2018). Random forest as a generic framework for predictive modeling of spatial and spatio-temporal variables. PeerJ, 6, e5518. doi:10.7717/peerj.5518. https://doi.org/10.7717/peerj.5518.
